Descriptive, not injunctive, social norms caused increases in mask wearing during the COVID-19 pandemic

### 2023-03-15 manuscript.tex
update manuscript

Research has investigated how social norms emerge in a population over time. In the long term, cultural evolutionary models show that injunctive social norms can be vertically transmitted across generations by imitation or teaching, or horizontally diffused from neighboring populations\textsuperscript{6}. However, less is known about how social norms arise endogenously within populations in the short term, including how descriptive norms arise and interact with injunctive norms to shape behavior. Recent work in behavioral economics suggests that social norms of public good provisioning develop in tandem with cooperative behavior through repeated interactions\textsuperscript{11}. But it remains unclear whether these findings generalize beyond the laboratory to real human populations.

With regards to normative influences on behavior, studies have demonstrated positive effects of descriptive norms on a variety of cooperative behaviors, including recycling\textsuperscript{12}, paying taxes\textsuperscript{13}, and sustainably reusing towels in hotels\textsuperscript{14}. However, these studies have two key aspects that limit their ability to assess the causal impact of norms, both of which we address in our current work. First, studies have not accounted for other potential non-social explanations for behavior, such as perceptions of the effectiveness of the behavior and personal beliefs that one should behave in a certain way. These non-social beliefs, labeled ``factual beliefs'' and ``personal normative beliefs''\textsuperscript{15}, often correlate with descriptive and injunctive norms, but they are fundamentally different because they can cause behavior separately from social expectations about what others do or think should be done. For example, willingness to recycle might be driven by perceptions that recycling has a positive impact on the environment and/or personal beliefs that recycling is the right thing to do, even if social norms actively discourage recycling (e.g., recycling is not a common or socially approved behavior). It is thus important to control for non-social beliefs in studies of social norms. Second, studies have tended to follow cross-sectional experimental designs in which social norm perceptions are manipulated by the researchers. However, social norms are not static: they change dynamically over time through processes of deliberation and interaction\textsuperscript{16}. An alternative but underutilized method of assessing causality between social norms and cooperative behavior, while retaining ecological validity, is to follow these variables over time amidst a real, unfolding social dilemma.

To understand how social norms emerge over time and shape cooperative behavior in a non-experimental setting, we focus on mask wearing in the United States during the COVID-19 pandemic. In April 2020, one month after the World Health Organization declared COVID-19 a global pandemic, mask wearing was officially recommended by the Centers for Disease Control and Prevention (CDC) to minimize the spread of the disease\textsuperscript{17}. Mask wearing has individual benefits, but the CDC also emphasized the collective benefits in reducing disease spread\textsuperscript{18}. Indeed, mask wearing posed a social dilemma to many individuals, in that it imposed personal costs (e.g., difficulty breathing, disrupted social interaction) for the benefit of the community (e.g., ``flattening the curve'' to protect at-risk individuals). Thus, the development of mask wearing during the COVID-19 pandemic enables us to study the emergence of social norms and their causal effects on cooperative behavior over a short timescale within a single population.

Recent research has found positive relationships between social norms and protective COVID-19 behaviors. In the United States, one study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure\textsuperscript{19}, and another study found that experimentally-induced descriptive norms increased mask wearing intentions\textsuperscript{20}. In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing\textsuperscript{21}. These studies are telling, but since they are cross-sectional or only minimally longitudinal, they do not have the temporal granularity to capture fluctuating changes in norm strength and adherence across the pandemic. Furthermore, several of the studies do not control for political ideology, which is important to account for since COVID-19 was highly politicized in the United States\textsuperscript{22}.

@@

To understand how mask wearing social norms emerged and fluctuated over the course of the COVID-19 pandemic, we first visualized the average descriptive trends of self-reported norm perceptions across the entire study duration. Figure \ref{fig:plotTimeline} plots self-reported mask wearing and perceptions of descriptive and injunctive mask wearing norms alongside relevant pandemic-related events in the United States, such as CDC public health recommendations and COVID-19 case numbers. These events were obtained from the CDC Museum's COVID-19 Timeline\textsuperscript{17}.

@@

In late 2020 and throughout 2021, we see several cross-lagged effects from perceived descriptive norms to future mask wearing. On four occasions, within-person increases in perceived descriptive norms predicted future within-person increases in mask wearing. According to recent effect size guidelines for cross-lagged panel models\textsuperscript{23}, the standardized beta coefficients for these cross-lagged effects were large (first wave, \(\beta\) = 0.17, 95\% CI {[}0.06 0.28{]}, \emph{p} = .002; second wave, \(\beta\) = 0.21, 95\% CI {[}0.08 0.34{]}, \emph{p} = .001; fourth wave, \(\beta\) = 0.15, 95\% CI {[}0.01 0.30{]}, \emph{p} = .041; fifth wave, \(\beta\) = 0.16, 95\% CI {[}0.02 0.29{]}, \emph{p} = .023). In 2022, these cross-lagged effects from descriptive norms to mask wearing diminished. We also find some evidence for a reciprocal effect, whereby within-person increases in mask wearing predicted future within-person increases in perceived descriptive norms. Moreover, several cross-lagged effects emerged between perceived descriptive and injunctive norms, demonstrating reciprocal within-person causal effects between these variables.

@@

Our finding that social norms and mask wearing are tightly coupled over time provides real-world support for experimental evidence that social norms and cooperative behavior develop synergistically within groups via processes of social interaction\textsuperscript{11}. The fact that these changes closely tracked the release of guidelines by the CDC supports the idea that institutions are part of the process by which culture and one's own behaviors are mutually constructed\textsuperscript{24}. Indeed, empirical work in cultural evolution suggests that formal institutions are critical for the emergence and rapid adoption of novel social norms\textsuperscript{25}. While new norms can and do emerge spontaneously in populations, the process is slow compared to institution-driven norm change, which, as our trends have shown, can unfold over measurement intervals as short as four to six weeks.

We found that descriptive norms, not injunctive norms, predicted future within-person increases in mask wearing, independent of the effects of non-social beliefs and political orientation. This finding is in line with previous evidence showing that perceptions of descriptive norms were positively related to other protective COVID-19 behaviors\textsuperscript{20,21}. There are several explanations for why descriptive norms have had these positive effects on protective COVID-19 behaviors like mask wearing. First, people may have followed descriptive norms to quickly coordinate their behavior with others during the pandemic. Descriptive norms are particularly useful for coordinating behavior during fast changing, threatening situations with a high degree of uncertainty, such as the COVID-19 pandemic\textsuperscript{26}. Second, people might have engaged in conditional cooperation, adapting their cooperation levels to the degree of cooperation in the population\textsuperscript{27}. Descriptive mask wearing norms provide evidence that others are cooperating, increasing the likelihood that individuals will themselves contribute to the public good by wearing masks. Third, the increased frequency of mask wearing in the population might have created a bandwagon effect\textsuperscript{28}, encouraging conformist copying. Under this view, people wear masks not to coordinate or cooperate, but simply to fit in with the crowd. Future research will be required to determine the motivations underlying adherence to descriptive norms during uncertain times.

@@

For the case of mask wearing in the United States during the COVID-19 pandemic, we have shown that social norms developed rapidly in the population and tracked ongoing changes in both recommendations from authorities and current levels of cooperative behavior. Moreover, we found that descriptive norms, rather than injunctive norms, were the main driver for future mask wearing. Importantly, this key finding slices two ways. Not only does it imply that high local levels of mask wearing encouraged future personal mask use, but it also implies that \emph{low} local levels of mask wearing \emph{discouraged} future personal mask use. This echoes recent reports of people in the United States not wanting to be ``singled out'' by being the only one wearing a mask in their community\textsuperscript{29}. Our work thus underscores the importance of consistent, visible community adherence for encouraging protective behaviors in response to global pandemics like COVID-19.

@@

To identify direct causal effects in our longitudinal analysis, we constructed a directed acyclic causal graph outlining the expected causal relationships between our variables (see Supplementary Figure \ref{fig:plotDAG}). In this causal model, we included two kinds of non-social beliefs highlighted by previous research\textsuperscript{15}: factual beliefs (i.e., beliefs about the effectiveness or consequences of mask wearing) and personal normative beliefs (i.e., personal beliefs about whether mask wearing is the right thing to do). These variables were included as potential mediators of the effects of descriptive and injunctive social norms on mask wearing. In addition, we also included political orientation as a common cause of all other variables. This is justified by evidence showing that mask wearing was heavily politicized in the United States during the pandemic\textsuperscript{22}. Given this causal graph, it is necessary to control for factual beliefs, personal normative beliefs, and political orientation in order to estimate the direct causal effects of descriptive and injunctive norms on mask wearing behavior over time.

@@

To analyze average trends in self-reported mask wearing and perceived social norms, we fitted several multilevel regression models. First, to determine whether mask wearing and social norms were coupled over time, we regressed mask wearing on perceived descriptive and injunctive norms separately, including random intercepts and slopes for participants and time points. Second, to analyze whether changes over time were related to recommendations from the CDC, we regressed mask wearing and perceived social norms onto a continuous time predictor. These models included random intercepts and slopes for participants, as well as change points aligning with changes in CDC mask wearing recommendations. We estimated these multilevel regression models using the \emph{lme4} R package\textsuperscript{30} and dealt with missing data via listwise deletion.

To quantify the within-person relationships between our variables over time, we fitted a random-intercept cross-lagged panel model to our longitudinal data\textsuperscript{31}. This structural equation model distinguishes between stable between-person trait levels and within-person fluctuations from trait levels. Positive cross-lagged effects from this model indicate that being above average on one variable at time \emph{t-1} predicts being above average in another variable at time \emph{t}. These models are considered the gold standard for identifying Granger causality in longitudinal datasets\textsuperscript{31,32}.

We estimated the random-intercept cross-lagged panel model using the \emph{lavaan} R package\textsuperscript{33}. In line with our directed acyclic graph (see Supplementary Figure \ref{fig:plotDAG}), we included three main variables (self-reported mask wearing, perceived descriptive norms, and perceived injunctive norms) and two time-variant control variables (factual beliefs and personal normative beliefs) in the model. For each of these variables, the model estimated a stable between-person trait level (random intercept) and time-specific within-person fluctuations from this trait level. We modeled autoregressive and cross-lagged effects between all five variables, and included political orientation as a time-invariant covariate. We restricted the analysis to the ten time points with available data for all five variables. Full information maximum likelihood estimation was used to deal with missing data.

All analyses were conducted in R v4.1.1\textsuperscript{34}. Visualizations were generated using the \emph{cowplot}\textsuperscript{35} and \emph{ggplot2}\textsuperscript{36} packages. The manuscript was reproducibly generated using the \emph{targets}\textsuperscript{37} and \emph{papaja}\textsuperscript{38} packages. All code and data are publicly available on GitHub\textsuperscript{39}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Titlestad2019}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Titlestad, K., Snijders, T. A. B., Durrheim, K., Quayle, M. \& Postmes, T. \href{https://doi.org/10.1016/j.jesp.2019.03.010}{The dynamic emergence of cooperative norms in a social dilemma}. \emph{Journal of Experimental Social Psychology} \textbf{84}, 103799 (2019).}

\leavevmode\vadjust pre{\hypertarget{ref-Nigbur2010}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Nigbur, D., Lyons, E. \& Uzzell, D. \href{https://doi.org/10.1348/014466609X449395}{Attitudes, norms, identity and environmental behaviour: Using an expanded theory of planned behaviour to predict participation in a kerbside recycling programme}. \emph{British Journal of Social Psychology} \textbf{49}, 259--284 (2010).}

\leavevmode\vadjust pre{\hypertarget{ref-Larkin2018}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{Larkin, C., Sanders, M., Andresen, I. \& Algate, F. \emph{Testing local descriptive norms and salience of enforcement action: A field experiment to increase tax collection}. (2018) doi:\href{https://doi.org/10.2139/ssrn.3167575}{10.2139/ssrn.3167575}.}

\leavevmode\vadjust pre{\hypertarget{ref-Goldstein2008}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Goldstein, N. J., Cialdini, R. B. \& Griskevicius, V. \href{https://doi.org/10.1086/586910}{A room with a viewpoint: Using social norms to motivate environmental conservation in hotels}. \emph{Journal of Consumer Research} \textbf{35}, 472--482 (2008).}

\leavevmode\vadjust pre{\hypertarget{ref-Bicchieri2014}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{Bicchieri, C., Lindemans, J. W. \& Jiang, T. \href{https://doi.org/10.3389/fpsyg.2014.01418}{A structured approach to a diagnostic of collective practices}. \emph{Frontiers in Psychology} \textbf{5}, 1418 (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-Smith2015}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{Smith, L. G. E., Thomas, E. F. \& McGarty, C. \href{https://doi.org/10.1111/pops.12180}{"We must be the change we want to see in the world": Integrating norms and identities through social interaction}. \emph{Political Psychology} \textbf{36}, 543--557 (2015).}

\leavevmode\vadjust pre{\hypertarget{ref-CDC2022}{}}%
\CSLLeftMargin{17. }%
\CSLRightInline{Centers for Disease Control and Prevention. \href{https://www.cdc.gov/museum/timeline/covid19.html}{{CDC Museum COVID-19 Timeline}}. (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-CDC2020}{}}%
\CSLLeftMargin{18. }%
\CSLRightInline{Centers for Disease Control and Prevention. \href{https://www.cdc.gov/media/releases/2020/p0714-americans-to-wear-masks.html}{{CDC calls on Americans to wear masks to prevent COVID-19 spread}}. (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Macy2021}{}}%
\CSLLeftMargin{19. }%
\CSLRightInline{Macy, J. T., Owens, C., Mullis, K. \& Middlestadt, S. E. \href{https://doi.org/10.3389/fpubh.2021.660813}{The role of self-efficacy and injunctive norms in helping older adults decide to stay home during the {COVID-19} pandemic}. \emph{Frontiers in Public Health} \textbf{9}, 660813 (2021).}

\leavevmode\vadjust pre{\hypertarget{ref-Bokemper2021}{}}%
\CSLLeftMargin{20. }%
\CSLRightInline{Bokemper, S. E. \emph{et al.} \href{https://doi.org/10.1371/journal.pone.0258282}{Experimental evidence that changing beliefs about mask efficacy and social norms increase mask wearing for {COVID-19} risk reduction: Results from the {United States} and {Italy}}. \emph{PLOS ONE} \textbf{16}, e0258282 (2021).}

\leavevmode\vadjust pre{\hypertarget{ref-Rudert2021}{}}%
\CSLLeftMargin{21. }%
\CSLRightInline{Rudert, S. C. \& Janke, S. \href{https://doi.org/10.1177/13684302211023562}{Following the crowd in times of crisis: Descriptive norms predict physical distancing, stockpiling, and prosocial behavior during the {COVID-19} pandemic}. \emph{Group Processes \& Intergroup Relations} \textbf{25}, 1819--1835 (2021).}

\leavevmode\vadjust pre{\hypertarget{ref-BaxterKing2022}{}}%
\CSLLeftMargin{22. }%
\CSLRightInline{Baxter-King, R., Brown, J. R., Enos, R. D., Naeim, A. \& Vavreck, L. \href{https://doi.org/10.1073/pnas.2116311119}{How local partisan context conditions prosocial behaviors: Mask wearing during {COVID-19}}. \emph{Proceedings of the National Academy of Sciences} \textbf{119}, e2116311119 (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-Orth2022}{}}%
\CSLLeftMargin{23. }%
\CSLRightInline{Orth, U. \emph{et al.} Effect size guidelines for cross-lagged effects. \emph{Psychological Methods} (2022) doi:\href{https://doi.org/10.1037/met0000499}{10.1037/met0000499}.}

\leavevmode\vadjust pre{\hypertarget{ref-Markus2010}{}}%
\CSLLeftMargin{24. }%
\CSLRightInline{Markus, H. R. \& Kitayama, S. \href{https://doi.org/10.1177/1745691610375557}{Cultures and selves: A cycle of mutual constitution}. \emph{Perspectives on Psychological Science} \textbf{5}, 420--430 (2010).}

\leavevmode\vadjust pre{\hypertarget{ref-Amato2018}{}}%
\CSLLeftMargin{25. }%
\CSLRightInline{Amato, R., Lacasa, L., Díaz-Guilera, A. \& Baronchelli, A. \href{https://doi.org/10.1073/pnas.1721059115}{The dynamics of norm change in the cultural evolution of language}. \emph{Proceedings of the National Academy of Sciences} \textbf{115}, 8260--8265 (2018).}

\leavevmode\vadjust pre{\hypertarget{ref-Gelfand2015}{}}%
\CSLLeftMargin{26. }%
\CSLRightInline{Gelfand, M. J. \& Harrington, J. R. \href{https://doi.org/10.1177/0022022115600796}{The motivational force of descriptive norms: For whom and when are descriptive norms most predictive of behavior?} \emph{Journal of Cross-Cultural Psychology} \textbf{46}, 1273--1278 (2015).}

\leavevmode\vadjust pre{\hypertarget{ref-Chaudhuri2011}{}}%
\CSLLeftMargin{27. }%
\CSLRightInline{Chaudhuri, A. \href{https://doi.org/10.1007/s10683-010-9257-1}{Sustaining cooperation in laboratory public goods experiments: A selective survey of the literature}. \emph{Experimental Economics} \textbf{14}, 47--83 (2011).}

>>>>

Research has investigated how social norms emerge in a population over time. In the long term, cultural evolutionary models show that injunctive social norms can be vertically transmitted across generations by imitation or teaching, or horizontally diffused from neighboring populations\textsuperscript{6}. However, less is known about how social norms arise endogenously within populations in the short term. While researchers have simulated the emergence of descriptive norms\textsuperscript{11,12}, this modelling work does not capture how descriptive norms develop alongside injunctive norms in real-world settings. Recent work in behavioral economics has also suggested that injunctive norms of public good provisioning develop in tandem with cooperative behavior through repeated interactions\textsuperscript{13}. But it remains unclear whether these findings generalize beyond the laboratory to real human populations.

With regards to normative influences on behavior, studies have demonstrated positive effects of descriptive norms on a variety of cooperative behaviors, including recycling\textsuperscript{14}, paying taxes\textsuperscript{15}, and sustainably reusing towels in hotels\textsuperscript{16}. However, these studies have two key aspects that limit their ability to assess the causal impact of norms, both of which we address in our current work. First, studies have not accounted for other potential non-social explanations for behavior, such as perceptions of the effectiveness of the behavior and personal beliefs that one should behave in a certain way. These non-social beliefs, labeled ``factual beliefs'' and ``personal normative beliefs''\textsuperscript{17}, often correlate with descriptive and injunctive norms, but they are fundamentally different because they can cause behavior separately from social expectations about what others do or think should be done. For example, willingness to recycle might be driven by perceptions that recycling has a positive impact on the environment and/or personal beliefs that recycling is the right thing to do, even if social norms actively discourage recycling (e.g., recycling is not a common or socially approved behavior). It is thus important to control for non-social beliefs in studies of social norms. Second, studies have tended to follow cross-sectional experimental designs in which social norm perceptions are manipulated by the researchers. However, social norms are not static: they change dynamically over time through processes of deliberation and interaction\textsuperscript{18}. An alternative but underutilized method of assessing causality between social norms and cooperative behavior, while retaining ecological validity, is to follow these variables over time amidst a real, unfolding social dilemma.

To understand how social norms emerge over time and shape cooperative behavior in a non-experimental setting, we focus on mask wearing in the United States during the COVID-19 pandemic. In April 2020, one month after the World Health Organization declared COVID-19 a global pandemic, mask wearing was officially recommended by the Centers for Disease Control and Prevention (CDC) to minimize the spread of the disease\textsuperscript{19}. Mask wearing has individual benefits, but the CDC also emphasized the collective benefits in reducing disease spread\textsuperscript{20}. Indeed, mask wearing posed a social dilemma to many individuals, in that it imposed personal costs (e.g., difficulty breathing, disrupted social interaction) for the benefit of the community (e.g., ``flattening the curve'' to protect at-risk individuals). Thus, the development of mask wearing during the COVID-19 pandemic enables us to study the emergence of social norms and their causal effects on cooperative behavior over a short timescale within a single population.

Recent research has found positive relationships between social norms and protective COVID-19 behaviors. In the United States, one study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure\textsuperscript{21}, and another study found that experimentally-induced descriptive norms increased mask wearing intentions\textsuperscript{22}. In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing\textsuperscript{23}. These studies are telling, but since they are cross-sectional or only minimally longitudinal, they do not have the temporal granularity to capture fluctuating changes in norm strength and adherence across the pandemic. Furthermore, several of the studies do not control for political ideology, which is important to account for since COVID-19 was highly politicized in the United States\textsuperscript{24}.

@@

To understand how mask wearing social norms emerged and fluctuated over the course of the COVID-19 pandemic, we first visualized the average descriptive trends of self-reported norm perceptions across the entire study duration. Figure \ref{fig:plotTimeline} plots self-reported mask wearing and perceptions of descriptive and injunctive mask wearing norms alongside relevant pandemic-related events in the United States, such as CDC public health recommendations and COVID-19 case numbers. These events were obtained from the CDC Museum's COVID-19 Timeline\textsuperscript{19}.

@@

In late 2020 and throughout 2021, we see several cross-lagged effects from perceived descriptive norms to future mask wearing. On four occasions, within-person increases in perceived descriptive norms predicted future within-person increases in mask wearing. According to recent effect size guidelines for cross-lagged panel models\textsuperscript{25}, the standardized beta coefficients for these cross-lagged effects were large (first wave, \(\beta\) = 0.17, 95\% CI {[}0.06 0.28{]}, \emph{p} = .002; second wave, \(\beta\) = 0.21, 95\% CI {[}0.08 0.34{]}, \emph{p} = .001; fourth wave, \(\beta\) = 0.15, 95\% CI {[}0.01 0.30{]}, \emph{p} = .041; fifth wave, \(\beta\) = 0.16, 95\% CI {[}0.02 0.29{]}, \emph{p} = .023). In 2022, these cross-lagged effects from descriptive norms to mask wearing diminished. We also find some evidence for a reciprocal effect, whereby within-person increases in mask wearing predicted future within-person increases in perceived descriptive norms. Moreover, several cross-lagged effects emerged between perceived descriptive and injunctive norms, demonstrating reciprocal within-person causal effects between these variables.

@@

Our finding that social norms and mask wearing are tightly coupled over time provides real-world support for experimental evidence that social norms and cooperative behavior develop synergistically within groups via processes of social interaction\textsuperscript{13}. The fact that these changes closely tracked the release of guidelines by the CDC supports the idea that institutions are part of the process by which culture and one's own behaviors are mutually constructed\textsuperscript{26}. Indeed, empirical work in cultural evolution suggests that formal institutions are critical for the emergence and rapid adoption of novel social norms\textsuperscript{27}. While new norms can and do emerge spontaneously in populations, the process is slow compared to institution-driven norm change, which, as our trends have shown, can unfold over measurement intervals as short as four to six weeks.

We found that descriptive norms, not injunctive norms, predicted future within-person increases in mask wearing, independent of the effects of non-social beliefs and political orientation. This finding is in line with previous evidence showing that perceptions of descriptive norms were positively related to other protective COVID-19 behaviors\textsuperscript{22,23}. There are several explanations for why descriptive norms have had these positive effects on protective COVID-19 behaviors like mask wearing. First, people may have followed descriptive norms to quickly coordinate their behavior with others during the pandemic. Descriptive norms are particularly useful for coordinating behavior during fast changing, threatening situations with a high degree of uncertainty, such as the COVID-19 pandemic\textsuperscript{28}. Second, people might have engaged in conditional cooperation, adapting their cooperation levels to the degree of cooperation in the population\textsuperscript{29}. Descriptive mask wearing norms provide evidence that others are cooperating, increasing the likelihood that individuals will themselves contribute to the public good by wearing masks. Third, the increased frequency of mask wearing in the population might have created a bandwagon effect\textsuperscript{30}, encouraging conformist copying. Under this view, people wear masks not to coordinate or cooperate, but simply to fit in with the crowd. Future research will be required to determine the motivations underlying adherence to descriptive norms during uncertain times.

@@

For the case of mask wearing in the United States during the COVID-19 pandemic, we have shown that social norms developed rapidly in the population and tracked ongoing changes in both recommendations from authorities and current levels of cooperative behavior. Moreover, we found that descriptive norms, rather than injunctive norms, were the main driver for future mask wearing. Importantly, this key finding slices two ways. Not only does it imply that high local levels of mask wearing encouraged future personal mask use, but it also implies that \emph{low} local levels of mask wearing \emph{discouraged} future personal mask use. This echoes recent reports of people in the United States not wanting to be ``singled out'' by being the only one wearing a mask in their community\textsuperscript{31}. Our work thus underscores the importance of consistent, visible community adherence for encouraging protective behaviors in response to global pandemics like COVID-19.

@@

To identify direct causal effects in our longitudinal analysis, we constructed a directed acyclic causal graph outlining the expected causal relationships between our variables (see Supplementary Figure \ref{fig:plotDAG}). In this causal model, we included two kinds of non-social beliefs highlighted by previous research\textsuperscript{17}: factual beliefs (i.e., beliefs about the effectiveness or consequences of mask wearing) and personal normative beliefs (i.e., personal beliefs about whether mask wearing is the right thing to do). These variables were included as potential mediators of the effects of descriptive and injunctive social norms on mask wearing. In addition, we also included political orientation as a common cause of all other variables. This is justified by evidence showing that mask wearing was heavily politicized in the United States during the pandemic\textsuperscript{24}. Given this causal graph, it is necessary to control for factual beliefs, personal normative beliefs, and political orientation in order to estimate the direct causal effects of descriptive and injunctive norms on mask wearing behavior over time.

@@

To analyze average trends in self-reported mask wearing and perceived social norms, we fitted several multilevel regression models. First, to determine whether mask wearing and social norms were coupled over time, we regressed mask wearing on perceived descriptive and injunctive norms separately, including random intercepts and slopes for participants and time points. Second, to analyze whether changes over time were related to recommendations from the CDC, we regressed mask wearing and perceived social norms onto a continuous time predictor. These models included random intercepts and slopes for participants, as well as change points aligning with changes in CDC mask wearing recommendations. We estimated these multilevel regression models using the \emph{lme4} R package\textsuperscript{32} and dealt with missing data via listwise deletion.

To quantify the within-person relationships between our variables over time, we fitted a random-intercept cross-lagged panel model to our longitudinal data\textsuperscript{33}. This structural equation model distinguishes between stable between-person trait levels and within-person fluctuations from trait levels. Positive cross-lagged effects from this model indicate that being above average on one variable at time \emph{t-1} predicts being above average in another variable at time \emph{t}. These models are considered the gold standard for identifying Granger causality in longitudinal datasets\textsuperscript{33,34}.

We estimated the random-intercept cross-lagged panel model using the \emph{lavaan} R package\textsuperscript{35}. In line with our directed acyclic graph (see Supplementary Figure \ref{fig:plotDAG}), we included three main variables (self-reported mask wearing, perceived descriptive norms, and perceived injunctive norms) and two time-variant control variables (factual beliefs and personal normative beliefs) in the model. For each of these variables, the model estimated a stable between-person trait level (random intercept) and time-specific within-person fluctuations from this trait level. We modeled autoregressive and cross-lagged effects between all five variables, and included political orientation as a time-invariant covariate. We restricted the analysis to the ten time points with available data for all five variables. Full information maximum likelihood estimation was used to deal with missing data.

All analyses were conducted in R v4.1.1\textsuperscript{36}. Visualizations were generated using the \emph{cowplot}\textsuperscript{37} and \emph{ggplot2}\textsuperscript{38} packages. The manuscript was reproducibly generated using the \emph{targets}\textsuperscript{39} and \emph{papaja}\textsuperscript{40} packages. All code and data are publicly available on GitHub\textsuperscript{41}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Muldoon2014a}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Muldoon, R., Lisciandra, C. \& Hartmann, S. Why are there descriptive norms? Because we looked for them. \emph{Synthese} \textbf{191}, 4409--4429 (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-Muldoon2014b}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Muldoon, R., Lisciandra, C., Bicchieri, C., Hartmann, S. \& Sprenger, J. On the emergence of descriptive norms. \emph{Politics, Philosophy \& Economics} \textbf{13}, 3--22 (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-Titlestad2019}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{Titlestad, K., Snijders, T. A. B., Durrheim, K., Quayle, M. \& Postmes, T. \href{https://doi.org/10.1016/j.jesp.2019.03.010}{The dynamic emergence of cooperative norms in a social dilemma}. \emph{Journal of Experimental Social Psychology} \textbf{84}, 103799 (2019).}

\leavevmode\vadjust pre{\hypertarget{ref-Nigbur2010}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Nigbur, D., Lyons, E. \& Uzzell, D. \href{https://doi.org/10.1348/014466609X449395}{Attitudes, norms, identity and environmental behaviour: Using an expanded theory of planned behaviour to predict participation in a kerbside recycling programme}. \emph{British Journal of Social Psychology} \textbf{49}, 259--284 (2010).}

\leavevmode\vadjust pre{\hypertarget{ref-Larkin2018}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{Larkin, C., Sanders, M., Andresen, I. \& Algate, F. \emph{Testing local descriptive norms and salience of enforcement action: A field experiment to increase tax collection}. (2018) doi:\href{https://doi.org/10.2139/ssrn.3167575}{10.2139/ssrn.3167575}.}

\leavevmode\vadjust pre{\hypertarget{ref-Goldstein2008}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{Goldstein, N. J., Cialdini, R. B. \& Griskevicius, V. \href{https://doi.org/10.1086/586910}{A room with a viewpoint: Using social norms to motivate environmental conservation in hotels}. \emph{Journal of Consumer Research} \textbf{35}, 472--482 (2008).}

\leavevmode\vadjust pre{\hypertarget{ref-Bicchieri2014}{}}%
\CSLLeftMargin{17. }%
\CSLRightInline{Bicchieri, C., Lindemans, J. W. \& Jiang, T. \href{https://doi.org/10.3389/fpsyg.2014.01418}{A structured approach to a diagnostic of collective practices}. \emph{Frontiers in Psychology} \textbf{5}, 1418 (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-Smith2015}{}}%
\CSLLeftMargin{18. }%
\CSLRightInline{Smith, L. G. E., Thomas, E. F. \& McGarty, C. \href{https://doi.org/10.1111/pops.12180}{"We must be the change we want to see in the world": Integrating norms and identities through social interaction}. \emph{Political Psychology} \textbf{36}, 543--557 (2015).}

\leavevmode\vadjust pre{\hypertarget{ref-CDC2022}{}}%
\CSLLeftMargin{19. }%
\CSLRightInline{Centers for Disease Control and Prevention. \href{https://www.cdc.gov/museum/timeline/covid19.html}{{CDC Museum COVID-19 Timeline}}. (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-CDC2020}{}}%
\CSLLeftMargin{20. }%
\CSLRightInline{Centers for Disease Control and Prevention. \href{https://www.cdc.gov/media/releases/2020/p0714-americans-to-wear-masks.html}{{CDC calls on Americans to wear masks to prevent COVID-19 spread}}. (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Macy2021}{}}%
\CSLLeftMargin{21. }%
\CSLRightInline{Macy, J. T., Owens, C., Mullis, K. \& Middlestadt, S. E. \href{https://doi.org/10.3389/fpubh.2021.660813}{The role of self-efficacy and injunctive norms in helping older adults decide to stay home during the {COVID-19} pandemic}. \emph{Frontiers in Public Health} \textbf{9}, 660813 (2021).}

\leavevmode\vadjust pre{\hypertarget{ref-Bokemper2021}{}}%
\CSLLeftMargin{22. }%
\CSLRightInline{Bokemper, S. E. \emph{et al.} \href{https://doi.org/10.1371/journal.pone.0258282}{Experimental evidence that changing beliefs about mask efficacy and social norms increase mask wearing for {COVID-19} risk reduction: Results from the {United States} and {Italy}}. \emph{PLOS ONE} \textbf{16}, e0258282 (2021).}

\leavevmode\vadjust pre{\hypertarget{ref-Rudert2021}{}}%
\CSLLeftMargin{23. }%
\CSLRightInline{Rudert, S. C. \& Janke, S. \href{https://doi.org/10.1177/13684302211023562}{Following the crowd in times of crisis: Descriptive norms predict physical distancing, stockpiling, and prosocial behavior during the {COVID-19} pandemic}. \emph{Group Processes \& Intergroup Relations} \textbf{25}, 1819--1835 (2021).}

\leavevmode\vadjust pre{\hypertarget{ref-BaxterKing2022}{}}%
\CSLLeftMargin{24. }%
\CSLRightInline{Baxter-King, R., Brown, J. R., Enos, R. D., Naeim, A. \& Vavreck, L. \href{https://doi.org/10.1073/pnas.2116311119}{How local partisan context conditions prosocial behaviors: Mask wearing during {COVID-19}}. \emph{Proceedings of the National Academy of Sciences} \textbf{119}, e2116311119 (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-Orth2022}{}}%
\CSLLeftMargin{25. }%
\CSLRightInline{Orth, U. \emph{et al.} Effect size guidelines for cross-lagged effects. \emph{Psychological Methods} (2022) doi:\href{https://doi.org/10.1037/met0000499}{10.1037/met0000499}.}

\leavevmode\vadjust pre{\hypertarget{ref-Markus2010}{}}%
\CSLLeftMargin{26. }%
\CSLRightInline{Markus, H. R. \& Kitayama, S. \href{https://doi.org/10.1177/1745691610375557}{Cultures and selves: A cycle of mutual constitution}. \emph{Perspectives on Psychological Science} \textbf{5}, 420--430 (2010).}

\leavevmode\vadjust pre{\hypertarget{ref-Amato2018}{}}%
\CSLLeftMargin{27. }%
\CSLRightInline{Amato, R., Lacasa, L., Díaz-Guilera, A. \& Baronchelli, A. \href{https://doi.org/10.1073/pnas.1721059115}{The dynamics of norm change in the cultural evolution of language}. \emph{Proceedings of the National Academy of Sciences} \textbf{115}, 8260--8265 (2018).}

\leavevmode\vadjust pre{\hypertarget{ref-Gelfand2015}{}}%
\CSLLeftMargin{28. }%
\CSLRightInline{Gelfand, M. J. \& Harrington, J. R. \href{https://doi.org/10.1177/0022022115600796}{The motivational force of descriptive norms: For whom and when are descriptive norms most predictive of behavior?} \emph{Journal of Cross-Cultural Psychology} \textbf{46}, 1273--1278 (2015).}
